A generalized definition of multidimensional item response theory parameters

### 2023-02-15 output/paper_draft/paper_M2PLM_parameters.Rmd
pedido git

Multidimensional item response theory (MIRT) models have been widely used since
their inception.<!--# Ref. needed? -->
The so called *Multidimensional 2-Parameter Logistic*
[M2PL, @mckinley_extension_1983] model is among the most important
and widely used ones.
This is partly due to its close relationship to the common factor model
[@mcdonald_test_1999], but also to the formal tractability of its formulation.<!--# Ref. needed? -->
It consists of a generalization of the 2-parameter logistic model
[@birnbaum_latent_1968] to more than one latent trait
[@mckinley_extension_1983].
Its usefulness lies in describing the variability of the population when
the correct response does not depend on a single trait;
more specifically, when a deficit in a certain trait
and a surplus in another one cancel out their effects. Because of this property,
it is sometimes referred to as a *compensatory* model.

@reckase_difficulty_1985 argues for a need to take dimensionality into account
when characterizing the statistical properties of multidimensional items.
As such, he outlines a general procedure for determining their difficulty.
Following this procedure, he also derives
a *multidimensional difficulty* parameter for M2PL items.
This parameter consists of two parts:
The first one is a scalar value representing the distance from
the origin of the multidimensional latent space to
the maximally discriminating locus of the item.
The second one, the direction of the vector from the origin to that same locus.
Following a similar rationale, @reckase_discriminating_1991 derive a
*multidimensional discrimination* (`r MDISC_PARAM`) parameter definition
for that model.

A crucial assumption leading to those formulations is that the axes of the
latent space are orthogonal
[@ackerman_using_1994; @ackerman_multidimensional_2005; see also
@reckase_difficulty_1985, p. 404 and Equation 7].
Of course, orthogonality is a property of a geometrical space instead of a
latent trait space.
Thus, nothing prevents a researcher from interpreting an abstraction such as
a latent trait space, no matter its structure, as an orthogonal space
--exploratory factor analysis software such as SPSS <!--# Reference needed? -->or
the R package *psych* <!--# Reference needed? -->graphically represent
factor loadings on orthogonal axes even after an oblique rotation.
Nevertheless, the assimilation of correlated latent traits to non-orthogonal
axes is prevalent in the common factor literature
[see, e.g., @harman_modern_1970].<!--# Thurstone's Multiple Factor Analysis, 1947? -->
This assumption is not so easily found in the MIRT literature though,
which mostly deals with uncorrelated dimensions
[see @reckase_multidimensional_2009 for a comprehensive treatment on the topic].
This fact is probably due to the different spotlights
these two theoretical programs put the focus on;
factor analysis aims at a clear interpretation of
the latent covariance structure, and thus focuses on finding
a structure matrix as simple as possible. MIRT literature, on the other hand,
aims at modeling responses to certain stimuli and estimating person scores,
being the model interpretation secondary in relevance.<!--# Reference needed? -->
Nevertheless, the confluence of the two theoretical approaches into a common
modern measurement theory [@mcdonald_test_1999] requires
the relaxation of traditional constrains on both approaches.
It is not uncommon to find recent examples of MIRT models where the
correlation matrix of the latent dimensions is freely estimated
[see, e.g., @reckase_multidimensional_2009, pp. 224-228]. Despite this, the
correspondence between correlated latent traits and orthogonal coordinate axes
is usually implicit in the MIRT literature without further consideration<!--# Exaples? -->
[see @zhang_theoretical_1999 for a notable exception].
Moreover, disregarding this geometrical interpretation may lead to
misconceptions that put the measurement and its conclusions into question.<!-- completar?? -->

A latent space where the traits are correlated is easier to understand as a
geometrical space with non-orthogonal axes,
often referred to as *general Cartesian coordinates* [@harman_modern_1970].
Such a space has the advantage that the cosines of the angles between two axes
can be interpreted as the correlation of the corresponding latent dimensions.
<!--# Reference needed? -->
<!-- Problem here: The latent space is not defined like that, according to
the results of section "Linear algebra of non-orthogonal coordinates". -->
The general case of MIRT model estimation implies estimating possibly
non-standard latent dimensions and non-null covariances among them
(a more realistic case than an identity matrix).
Assuming orthonormality among traits is usually unrealistic, be it in the
cognitive [e.g., @reckase_multidimensional_2009] or the non-cognitive domain<!--# Reference to example? -->.
Moreover, certain applications impose the estimation of
non-identity latent covariance matrices (e.g. multigroup equating) even when<!-- # Reference needed? -->
interpreting the latent dimensions as substantive constructs is not a goal.
As we will argue, the usual definitions of the multidimensional difficulty and
discrimination parameters do not generalize to the non-orthonormal geometry
necessary to accurately represent correlated latent spaces.

@@

The main objective of this paper is thus to provide a generalization of
the MIRT parameters in two aspects: on one hand, in an oblique coordinate system
that can be soundly applied to the correlated trait case; on the other, to
items that yield monotonous response probabilities, but in the broader sense of
unchanging monotony. Additionally, we aim at providing a method for
graphically representing these items in the fashion of @ackerman_creating_1994,
but taking the non-orthogonality of the coordinate axes into account.
In order to pursue these objectives, we first define the generalization of
the M2PL model and describe its scope of application.
Then, we follow the method outlined by @reckase_difficulty_1985 and
@reckase_discriminating_1991 to obtain
the formulation of the multidimensional parameters.
We will introduce two versions of these indices,
showing that either the original or one of the new ones is obtained
as a function of how we define and interpret the latent space.
Next we discuss the properties of the indices,
propose a graphical representation method,
and present two examples of application of our results.
We round up with a discussion about
the applicability and generalization of our results.

@@

where $`r DISCR_VECTOR`$ is an $`r N_DIMS`$-dimensional vector of discrimination
parameters, and $`r INTERCEPT_PARAM`$ an intercept parameter related to the
location of the item in the latent space.

@@

Let's consider the standard basis $`r LS_STD_BASIS_EQ`$ in $`r LATENT_SPACE`$.
We can write all $`r TRAIT_VEC_IN_LS`$ as a linear combination of each basis,
$`r LS_BASIS`$ or $`r LS_STD_BASIS`$.
Therefore, we have both representations: $`r TRAIT_VECTOR_DEF_BASIS_EQ`$, and
$`r TRAIT_VECTOR_DEF_STD_BASIS_EQ`$, where $`r TRAIT_COORDS_EQ`$ and
$`r TRAIT_COORDS_STD_EQ`$.

Let $`r BASIS_CHANGE_DEF`$ be the change of basis matrix between the two bases;
then, we have that

@@

The $`r DIM_INDEX`$-th column of $`r TRANSFORM_MATRIX`$ is given by the
coordinates of $`r BASIS_VECTOR_ANY`$ in basis $`r LS_STD_BASIS`$.<!--# Does this need to be justified? -->
If $`r BASIS_VECTOR_ANY_TRANSF_EQ`$, then

@@

With the standard basis in $`r LATENT_SPACE`$ we have the usual scalar product
(i.e., dot product).
With basis $`r LS_BASIS`$ in $`r LATENT_SPACE`$ we define an inner product
$`r LS_BASIS_INNER_PROD`$, such that the norm of $`r TRAIT_VECTOR`$ is
invariant, i.e., $`r TRAIT_NORM_EQ`$. Therefore,

@@

Thus, we must define the inner product $`r LS_BASIS_INNER_PROD`$ with Gram
matrix $`r INNER_PROD_MAT_EQ`$. That is, $`r INNER_PROD_EQ`$, which keeps the
norm invariant.
The Gram matrix has $`r INNER_PROD_MAT_ELEMENT_EQ`$ as element. As
$`r TRANSFORM_MATRIX`$ (and consequently $`r TRANSFORM_MATRIX_TRANSP`$) is
invertible, so $`r INNER_PROD_MATRIX`$ is, and $`r INNER_PROD_MATRIX_INV_EQ`$.

@@

Then, we have that $`r BASIS_CHANGE_ORG_ALT_EQ`$.

@@

As Equation \@ref(eq:M2PL-formula-polar) shows, the inner product matrix is
involved in the expression of the model in polar coordinates.
Importantly, this shows that @reckase_difficulty_1985's Equation 3 (p. 403)
already assumes the orthonormality of the latent space.

@@

We are thus considering the direction cosines with respect to the orthonormal
axes, so we can just apply @reckase_difficulty_1985's results, obtaining

@@

Applying Proposition 1,

@@

with $`r DIAG_MATRIX_INNER_PROD_INV`$ a diagonal matrix where
$`r MATRIX_INNER_PROD_INV_EL_DIAG_EQ`$.

>>>>

Multidimensional item response theory (MIRT) models have been widely used since
their inception.<!--# Ref. needed? -->
The so called *Multidimensional 2-Parameter Logistic*
[M2PL, @mckinley_extension_1983] model is one of the most important and
widely used multidimensional models,
partly due to its close relationship to the common factor model
[@mcdonald_test_1999], but also to the formal tractability of its formulation.<!--# Ref. needed? -->
The M2PL model is a generalization of the 2-parameter logistic model
[@birnbaum_latent_1968] to more than one latent trait
[@mckinley_extension_1983].
Its usefulness lies in describing the variability of the population when
the correct response does not depend on a single trait;
more specifically, when a deficit in a certain trait
and a surplus in another one cancel out their effects --thus it is sometimes
referred to as a *compensatory* model.

@reckase_difficulty_1985 argues in favor of a need to characterize
the statistical properties of items taking dimensionality into account.
As such, he outlines a general procedure for determining
the difficulty of multidimensional items.
Following this procedure, he also derives
a *multidimensional difficulty* parameter (MID) for M2PL items.
This parameter consists of two parts:
The first one is a scalar value representing the distance from
the origin of the multidimensional latent space to
the maximally discriminating point of the item.
The second one, the direction of the vector from the origin to that same point.
Following a similar rationale, @reckase_discriminating_1991 define
a *multidimensional discrimination* (MDISC) parameter definition for that model.

A crucial assumption leading to those formulations is that the axes of the
latent space are orthogonal
[@ackerman_using_1994; @ackerman_multidimensional_2005; see also @reckase_difficulty_1985, p. 404 and Equation 7].
Of course, orthogonality is a property of a geometrical space instead of a
latent trait space.
Thus, nothing prevents a researcher from interpreting an abstraction such as
a latent trait space, no matter its structure, as an orthogonal space
--exploratory factor analysis software such as SPSS <!--# Reference needed? -->or
the R package *psych* <!--# Reference needed? -->graphically represent
factor loadings on orthogonal axes, even after an oblique rotation.
Nevertheless, the assimilation of correlated latent traits to non-orthogonal
axes is prevalent in the common factor literature
[see, e.g., @harman_modern_1970].<!--# Thurstone's Multiple Factor Analysis, 1947? -->
This assumption is not so easily found in the MIRT literature though,
which mostly deals with uncorrelated dimensions
[see @reckase_multidimensional_2009 for a comprehensive treatment on the topic].
This fact is probably due to the different spotlights
the two theoretical programs put the focus on;
Factor analysis aims at a clear interpretation of
the latent covariance structure, and thus focuses on finding
a structure matrix as simple as possible. MIRT literature, on the other hand,
aims at modeling responses to certain stimuli and estimating person scores,
being the model interpretation secondary in relevance.<!--# Reference needed? -->
Nevertheless, the confluence of the two theoretical approaches into a common
modern measurement theory [@mcdonald_test_1999] imposes
the relaxation of traditional constrains on both approaches.
It is not uncommon to find recent examples of MIRT models where the
correlation matrix of the latent dimensions is freely estimated
[see, e.g., @reckase_multidimensional_2009, pp. 224-228]. Despite this, the
correspondence between correlated latent traits and orthogonal coordinate axes
is usually implicit in the MIRT literature without further consideration<!--# Exaples? -->
[see @zhang_theoretical_1999 for a notable exception].
What is more, disregarding this geometrical interpretation may lead to
misconceptions that put the measurement and its conclusions into question.<!-- completar?? -->

A latent space where the traits are correlated is easier to represent in
a geometrical space with non-orthogonal axes,
often referred to as *general Cartesian coordinates* [@harman_modern_1970].
Such a space has the advantage that the cosines of the angles between two axes
can be interpreted as the correlation of the corresponding latent dimensions.
<!--# Reference needed? -->
<!-- Problem here: The latent space is not defined like that, according to
the results of section "Linear algebra of non-orthogonal coordinates". -->
The general case of MIRT model estimation implies estimating possibly
non-standard latent dimensions and non-null covariances among them,
a more realistic case than an identity matrix.
Assuming orthogonality among traits is usually unrealistic, be it in the
cognitive [e.g., @reckase_multidimensional_2009] or the non-cognitive domain<!--# Reference to example? -->.
Moreover, certain applications impose the estimation of
non-identity latent covariance matrices (e.g. multigroup equating) even when<!-- # Reference needed? -->
interpreting the latent dimensions as substantive constructs is not a goal.
As we will argue, the usual definitions of the
MID and MDISC parameters do not generalize to the non-orthogonal geometry used
to represent correlated latent spaces.

@@

The main objective of this paper is thus to provide a generalization of
the MIRT parameters in two aspects: on one hand, in an oblique coordinate system
that can be soundly applied to the correlated trait case; on the other, to
items that yield monotonous response probabilities, but in the broader sense of
unchanging monotony. Additionally, we aim at providing a method for
graphically representing these items in the fashion of @ackerman_creating_1994,
but taking into account the non-orthogonality of the coordinate axes.
In order to pursue these objectives, we first define the generalization of
the M2PL model and describe its scope of application.
Then, we follow the method outlined by @reckase_difficulty_1985 and
@reckase_discriminating_1991 to obtain
the formulation of the multidimensional parameters.
We will introduce two versions of these indices,
showing that either the original or one of the new ones is obtained
as a function of how we define and interpret the latent space.
Next we discuss the properties of the indices,
propose a graphical representation method,
and present an example of computation<!-- # TODO: Pending -->
and representation of the items based on these parameters.
We round up with a discussion about
the applicability and generalization of our results.

@@

where $`r DISCR_VECTOR`$ is a $`r N_DIMS`$-dimensional vector of discrimination
parameters, and $`r INTERCEPT_PARAM`$ an intercept parameter related to the
location of the item in the latent space.

@@

Let's consider the standard basis $`r LS_STD_BASIS_EQ`$ in $`r LATENT_SPACE`$.
We can write all $`r TRAIT_VEC_IN_LS`$ as a linear combination of each baseis.
Therefore, we have both representations: $`r TRAIT_VECTOR_DEF_BASIS_EQ`$, and
$`r TRAIT_VECTOR_DEF_STD_BASIS_EQ`$, where $`r TRAIT_COORDS_EQ`$ and
$`r TRAIT_COORDS_STD_EQ`$.

Let $`r BASIS_CHANGE_DEF`$ be the change of basis matrix; then, we have that

@@

The $`r DIM_INDEX`$-th column of $`r TRANSFORM_MATRIX`$ is given by the
coordinates of $`r BASIS_VECTOR_ANY`$ in basis $`r LS_STD_BASIS`$.
If $`r BASIS_VECTOR_ANY_TRANSF_EQ`$, then

@@

With the standard basis in $`r LATENT_SPACE`$ we have the usual scalar product.
With basis $`r LS_BASIS`$ in $`r LATENT_SPACE`$ we define an inner product
$`r LS_BASIS_INNER_PROD`$, such that the norm of $`r TRAIT_VECTOR`$ is
invariant, i.e., $`r TRAIT_NORM_EQ`$. Therefore,

@@

Thus, we must define the inner product $`r LS_BASIS_INNER_PROD`$ with Gram
matrix $`r INNER_PROD_MAT_EQ`$. That is, $`r INNER_PROD_EQ`$, which keeps the
norm invariant.

The Gram matrix has $`r INNER_PROD_MAT_ELEMENT_EQ`$ as element. As
$`r TRANSFORM_MATRIX`$ (and consequently $`r TRANSFORM_MATRIX_TRANSP`$) is
invertible, so $`r INNER_PROD_MATRIX`$ is, and $`r INNER_PROD_MATRIX_INV_EQ`$.

@@

Then, we have that $`r BASIS_CHANGE_ORG_ALT_EQ`$. (ERROR aca)

@@

As Equation \@ref(eq:M2PL-formula-polar) shows, the inner product matrix is
involved in the expression of the model in polar coordinates.
Importantly, this shows that @reckase_difficulty_1985's Equation 3 (p. 403)
already assumes the orthogonality of the latent space.

@@

We are thus considering the direction cosines with respect to the orthogonal
axes, so we can just apply @reckase_difficulty_1985's results, obtaining

@@

Applying propositon 1,

@@

with $`r DIAG_MATRIX_INNER_PROD_INV`$ the diagonal matrix with
the same diagonal as $`r INNER_PROD_MATRIX_INV`$.
